---
title: "AI-Assisted Workflow for postlink"
author: "Isabella Liu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{AI-Assisted Workflow for postlink}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE,
  message  = FALSE,
  warning  = FALSE
)
```

## Overview

The **postlink** package provides two AI-assisted functions for interpreting
and reporting post-linkage data analyses:

| Function | Purpose |
|---|---|
| `explain()` | Plain-language interpretation of one or more fitted models |
| `generate_report()` | Structured analysis report (Data / Methods / Results / References) |
| `postlink_set_key()` | Store an API key in `~/.Renviron` using base R |
| `postlink_set_provider()` | Choose the LLM backend (Anthropic, OpenAI, Gemini, Ollama) |

Both `explain()` and `generate_report()` are **AI-agnostic**: they support
multiple LLM providers and automatically detect which to use based on the API
key available in the environment.

---

## 1. Setup and Configuration

### 1.1 Install dependencies

```{r install}
install.packages(c("httr2", "rmarkdown"))
devtools::install_github("postlink-group/postlink")
```

### 1.2 Configure the LLM backend

The `provider`, `api_key`, and `model` are parameters of `explain()` and
`generate_report()` — no separate setup functions are needed.

**Option A: Pass the key directly (quickest for testing)**

```{r key-direct}
library(postlink)

explain(fit_adj, api_key = "sk-ant-your-key-here")
generate_report(fit_adj, api_key = "sk-ant-your-key-here")
```

**Option B: Set an environment variable (recommended for regular use)**

Set once per session, or add to `~/.Renviron` so it loads automatically:

```{r key-env}
# Set for the current session
Sys.setenv(ANTHROPIC_API_KEY = "sk-ant-your-key-here")

# Or add to ~/.Renviron for all future sessions (use a text editor):
# ANTHROPIC_API_KEY=sk-ant-your-key-here

# Then just call without api_key argument:
explain(fit_adj)
```

**Supported providers:**

```{r providers}
# Anthropic Claude (default — used if ANTHROPIC_API_KEY is set)
explain(fit_adj, provider = "anthropic", api_key = "sk-ant-...")

# OpenAI
explain(fit_adj, provider = "openai", api_key = "sk-...")

# Google Gemini
explain(fit_adj, provider = "gemini", api_key = "AIza...")

# Local Ollama (no key needed — make sure Ollama is running: ollama serve)
explain(fit_adj, provider = "ollama", model = "llama3")
```

The provider is auto-detected from whichever environment variable is set.
Pass `provider` explicitly to override.

### 1.3 Load the package and data

```{r load}
library(postlink)

# Full LIFE-M dataset (~156k linked Ohio birth and death certificate records)
url        <- "https://github.com/postlink-group/postlink/releases/download/v0.1.0/lifem.rds"
lifem_full <- readRDS(url(url, "rb"))

cat("Records:", nrow(lifem_full), "\n")
cat("Columns:", paste(names(lifem_full), collapse = ", "), "\n")
```

---

## 2. Fit the Post-Linkage Models

Following the structure of the
[LIFE-M longevity analysis vignette](https://postlink-group.github.io/postlink/articles/longevity-analysis.html),
we fit three models for comparison:

- **Naive**: standard GLM, no linkage error correction
- **Hand-Linked Only**: GLM restricted to hand-verified matches (~1.4% of records)
- **Adjusted**: postlink EM mixture model correcting for linkage error

### 2.1 Naive model

```{r fit-naive}
fit_naive <- glm(
  age_at_death ~ poly(unit_yob, 3, raw = TRUE),
  data   = lifem_full,
  family = gaussian()
)
```

### 2.2 Hand-linked only model

```{r fit-hl}
lifem_hl <- lifem_full[lifem_full$hndlnk, ]
cat("Hand-linked records:", nrow(lifem_hl), "\n")  # ~2,159

fit_hl <- glm(
  age_at_death ~ poly(unit_yob, 3, raw = TRUE),
  data   = lifem_hl,
  family = gaussian()
)
```

### 2.3 Adjusted model (postlink EM mixture)

```{r fit-adj}
adj <- adjMixture(
  linked.data  = lifem_full,
  m.formula    = ~ commf + comml,   # name commonness covariates
  m.rate       = 0.05,              # mismatch rate upper bound
  safe.matches = hndlnk             # hand-linked records fixed at P(match) = 1
)

fit_adj <- plglm(
  age_at_death ~ poly(unit_yob, 3, raw = TRUE),
  family     = "gaussian",
  adjustment = adj
)
```

---

## 3. explain()

`explain()` accepts either a single model or a **named list of models**.
When a list is provided, the LLM automatically compares results across all
models and explains what the differences reveal about the impact of linkage
error correction.

### 3.1 Single model

```{r explain-single}
explain(
  fit_adj,
  context = "This study uses LIFE-M linked Ohio birth and death certificates
             (birth cohorts 1883-1906) to examine the relationship between
             year of birth and age at death. Records were linked using first
             and last name commonness scores; expected mismatch rate ~5%."
)
```

**Example output:**

```
--- postlink: explain() ---

The analysis used an EM Mixture Model (Slawski et al., 2025) to correct for
record linkage errors, explicitly modelling each linked record pair as either a
true match or a false link and iterating to recover regression coefficients
applicable to correctly matched records only; with a mean posterior correct match
probability of 0.90 and successful EM convergence, the implied ~10% mismatch
rate has been accounted for in the estimates reported here. The model regresses
the outcome on a cubic polynomial in year of birth (unit_yob), and all four
coefficients are highly statistically significant, each with |z| far exceeding 2.
The large and alternating signs of the three polynomial terms (linear: −51.17,
quadratic: +140.54, cubic: −76.92) together describe a non-linear, S-shaped or
wave-like relationship between birth year and the outcome, meaning the outcome
does not change at a constant rate across birth cohorts but instead accelerates
and decelerates at different points in the birth-year range. The intercept of
58.63 represents the model's baseline fitted value at the origin of the
polynomial, though its standalone interpretation depends on how unit_yob is
scaled. These corrected estimates reflect associations between birth year and the
outcome among true record matches, and should not be interpreted as causal
without additional substantive justification.

[AI-generated. Verify before use.]
```

### 3.2 Multi-model comparison

When `fit` is a named list, comparison mode is triggered automatically.

```{r explain-multi}
explain(
  list(
    "Naive (All Records)" = fit_naive,
    "Hand-Linked Only"    = fit_hl,
    "Adjusted (postlink)" = fit_adj
  ),
  context = "This study uses LIFE-M linked Ohio birth and death certificates
             (birth cohorts 1883-1906) to examine the relationship between
             year of birth and age at death."
)
```

**Example output:**

```
--- postlink: explain() — Model Comparison ---

The Naive model uses the full probabilistically linked dataset (n = 156,453)
without any correction for linkage errors, while the Hand-Linked model applies
the same unadjusted approach to a small gold-standard subset (n = 2,159)
verified through manual linkage, and the Adjusted model applies an EM mixture
model to the full linked dataset to correct for linkage error contamination.
Comparing coefficients, the Naive and Hand-Linked models produce broadly similar
estimates — for example, the cubic term is −72.94 and −59.92 respectively —
though the Hand-Linked estimates carry substantially larger standard errors
reflecting its much smaller sample size. The Adjusted model consistently produces
coefficients of greater magnitude than the Naive model across all polynomial terms
(e.g., the linear term shifts from −46.70 to −51.17, the quadratic from 130.41
to 140.54, and the cubic from −72.94 to −76.92), while simultaneously achieving
smaller standard errors than either unadjusted model. This pattern is
characteristic of linkage error correction: false matches introduce outcome
misclassification that attenuates associations toward the null, so the EM
adjustment recovers effect estimates closer to the true underlying relationship
while the larger full-sample size preserves precision. The Hand-Linked model
serves as a useful qualitative benchmark, and its coefficients trend in the same
direction as the Adjusted model relative to the Naive, lending credibility to the
correction.

[AI-generated. Verify before use.]
```

---

## 4. generate_report()

`generate_report()` also accepts a single model or a named list. When a list
is provided, a **Model Comparison** table is automatically included in the
Results section.

### 4.1 Single model report

```{r report-single}
generate_report(fit_adj, output_format = "text")
```

**Example output (with list of three models):**

```
postlink Analysis Report
========================================
August 30, 2026

## Data

The analysis used 156,453 records. The outcome variable is **age_at_death**.
Model formula: `age_at_death ~ poly(unit_yob, 3, raw = TRUE)`.

## Methods

Linkage error was accounted for using **postlink** (R package, v0.1.0; Bukke et
al., 2026). The outcome (age_at_death) was modelled using Gaussian (linear)
regression with formula `age_at_death ~ poly(unit_yob, 3, raw = TRUE)`. Linkage
error correction used an EM-based two-component mixture model (Slawski et al.,
2025). Each record pair is modelled as either a correct match (C = 1) or a false
link (C = 0). The EM algorithm alternates between estimating posterior correct
match probabilities (E-step) and updating the regression parameters for true
matches (M-step). The EM algorithm converged. The mean posterior correct match
probability was 0.9 (65.6% of records exceeded 0.9), implying an estimated
mismatch rate of approximately 10%.

### Mismatch Model Coefficients

| Term        | 95% CI Lower | 95% CI Upper |
|-------------|--------------|--------------|
| (Intercept) |  4.9355      |  6.7751      |
| commf       |  0.3946      |  2.5015      |
| comml       | -7.7461      | -6.4727      |

## Results

> The narrative below was drafted by an AI language model using exact numerical
> outputs from the model. Review and edit before submitting or publishing.

The EM mixture model, which down-weights false links across the full linked sample
(n = 156,453), yielded a significant cubic relationship between year of birth and
the outcome across all polynomial terms (all p < 0.05). The adjusted intercept
was 58.63 (95% CI: 58.34, 58.91), with linear, quadratic, and cubic year-of-birth
coefficients of −51.17 (95% CI: −54.20, −48.13), 140.54 (95% CI: 132.95,
148.12), and −76.92 (95% CI: −82.01, −71.83), respectively. Compared to the
naive unadjusted estimates, the EM-corrected coefficients show meaningfully
amplified polynomial terms — the linear term shifted from −46.70 to −51.17, the
quadratic from 130.41 to 140.54, and the cubic from −72.94 to −76.92 —
indicating that false links in the naive model attenuated the estimated trajectory
of the outcome across birth cohorts. The adjusted estimates align more closely
with those from the hand-linked validation sample (n = 2,159) in directional
pattern, demonstrating the value of the EM adjustment in recovering signal
obscured by linkage error.

### Model Comparison

| Term | Naive | Hand-Linked | Adjusted |
|------|-------|-------------|----------|
| (Intercept) | 58.4994 (0.2149) | 57.6558 (1.2603) | 58.6292 (0.1451) |
| poly(unit_yob, 3, raw = TRUE)1 | -46.7007 (1.7897) | -44.2281 (11.5935) | -51.1659 (1.5498) |
| poly(unit_yob, 3, raw = TRUE)2 | 130.4127 (3.9923) | 118.5657 (27.9187) | 140.5362 (3.8685) |
| poly(unit_yob, 3, raw = TRUE)3 | -72.9382 (2.5188) | -59.9173 (18.4589) | -76.9233 (2.5964) |
| n | 156,453 | 2,159 | 156,453 |

### Adjusted Model — Coefficient Table

| Term | Estimate | SE | z | p-value | 95% CI Lower | 95% CI Upper |
|------|----------|----|---|---------|--------------|--------------|
| (Intercept) | 58.6292 | 0.1451 | 404.061 | 0 * | 58.3448 | 58.9136 |
| poly(unit_yob, 3, raw = TRUE)1 | -51.1659 | 1.5498 | -33.015 | 0 * | -54.2034 | -48.1284 |
| poly(unit_yob, 3, raw = TRUE)2 | 140.5362 | 3.8685 | 36.328 | 0 * | 132.9541 | 148.1183 |
| poly(unit_yob, 3, raw = TRUE)3 | -76.9233 | 2.5964 | -29.627 | 0 * | -82.0122 | -71.8345 |

* p < 0.05

## References

- postlink R package (v0.1.0). Bukke, P., Kamat, G., Cui, J., Gutman, R., &
  Slawski, M. https://postlink-group.github.io/postlink/
- Slawski, M., West, B. T., Bukke, P., Wang, Z., Diao, G., & Ben-David, E.
  (2025). A general framework for regression with mismatched data based on
  mixture modelling. Journal of the Royal Statistical Society Series A,
  188(3), 896-919.
```

### 4.2 Full report with comparison

This is the recommended usage. It matches the LIFE-M vignette structure.

```{r report-full, eval=FALSE}
generate_report(
  list(
    "Naive (All Records)" = fit_naive,
    "Hand-Linked Only"    = fit_hl,
    "Adjusted"            = fit_adj
  ),
  output_file   = "articles/lifem_ai_report.html",
  output_format = "html",
  context = list(
    study            = "This study examines the relationship between year of
                        birth and age at death using linked vital records from
                        the LIFE-M project.",
    dataset          = "LIFE-M Ohio birth and death certificates, 1883-1906",
    n_total          = 156453,
    outcome_label    = "age at death (years)",
    predictor_label  = "year of birth",
    linkage_description = "Records were linked using first and last name
                           commonness scores. Hand-linked records (~1.4%)
                           were treated as confirmed correct matches.
                           Expected mismatch rate among machine-linked
                           records is approximately 5%."
  )
)
```

> The generated HTML report has been pre-built and is stored in the repository.
> **[View the generated report →](articles/lifem_ai_report.html)**

> To regenerate it yourself (requires an API key):
> ```r
> # Run the code block above with eval=TRUE
> ```

### 4.3 R Markdown output

```{r report-rmd}
generate_report(
  list(
    "Naive"       = fit_naive,
    "Hand-Linked" = fit_hl,
    "Adjusted"    = fit_adj
  ),
  output_file   = "lifem_report.Rmd",
  output_format = "rmd",
  render        = TRUE
)
```

---

## 5. Report Structure

The generated report contains four sections:

| Section | Content | Source |
|---|---|---|
| **Data** | Sample size, outcome, predictor | `fit` + `context` |
| **Methods** | Adjustment method details: algorithm, covariates, constraints, convergence | `fit` directly — no AI |
| **Results** | Comparison table (if list provided) + coefficient table + AI narrative | Tables from `fit`; narrative from AI |
| **References** | postlink package + method paper | Auto-populated by method |

---

## 6. Supported LLM Providers

| Provider | Environment Variable | Default Model | Usage |
|---|---|---|---|
| Anthropic (default) | `ANTHROPIC_API_KEY` | `claude-sonnet-4-6` | `explain(fit, provider="anthropic")` |
| OpenAI | `OPENAI_API_KEY` | `gpt-4o` | `explain(fit, provider="openai")` |
| Google Gemini | `GEMINI_API_KEY` | `gemini-1.5-flash` | `explain(fit, provider="gemini")` |
| Ollama (local) | none | `llama3` | `explain(fit, provider="ollama", model="llama3")` |

The provider is auto-detected from whichever environment variable is set.
Pass `provider` explicitly to override, or pass `api_key` directly to skip
environment variable lookup entirely.

---

## 7. Notes

- **No raw data sent**: Only numerical summaries (coefficients, SEs, n, method
  details) are sent to the LLM — never the raw linked dataset.

- **Always review AI output**: The LLM is instructed to use only the exact
  numbers provided, but all generated text should be reviewed before
  submitting or publishing.

- **Optional dependency**: `httr2` is listed under `Suggests` in the package
  `DESCRIPTION`. If not installed, a clear error message prompts installation.

- **Current limitations**: The Bayesian method (`adjMixBayes`) has not been
  fully tested. Cox regression and contingency table models have not been
  validated yet with these functions.
